---
title: "Phylogenetic Tree Construction with FastTree2"
author: "Jordan"
date: "2025-03-31"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goals 

The goal of this script is to make a phylogenetic tree so we can use it for phylogenetic community analyses like phylogenetic Hill Numbers and the UniFrac beta-diversity measures. 

1. Load in pre-processed phyloseq object. 
2. Create ASV fasta file from the phyloseq object.
3. Align the 16S Sequences from our fasta file using the software [MAFFT](https://academic.oup.com/mbe/article/30/4/772/1073398).  
4. Create a maximum likelihood tree using the software [FastTree2](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0009490). 
5. Write out the tree file for use in the next step of the analyses. 

## Inputs 
1. Our previously pre-processed phyloseq object that has chloroplasts, mitochondria, reverse complements, and negative controls removed, which is located in `data/03_PreProcessing/raw_preprocessed_physeq`

## Outputs 
1. The **aligned 16S rRNA gene sequences**, which is the input to the phylogenetic tree. We will save our alignment to `data/04_PhylogeneticTree/MAAFT_aligned_ASV.fasta`
2. The **calculated phylogenetic tree file**, which is the input to the phylogenetic tree. We will save our alignment to`data/04_PhylogeneticTree/ASVs_unrooted.tree`

# Why Phylogenetic Trees? 

In microbial community analyses, **phylogenetic trees** provide critical evolutionary context for interpreting the relationships among organisms. Unlike simple taxonomic classification, phylogenetic trees capture the shared ancestry and divergence between microbial sequences, which is especially useful when taxonomy is uncertain or incomplete. Many downstream ecological metrics—such as UniFrac dissimilarity and the Phylogenetic Hill Numbers—rely on an accurate tree to quantify differences in community composition or diversity while accounting for how closely related the organisms are. As such, building a phylogenetic tree from sequence data (*e.g.,* 16S rRNA gene sequences) is a foundational step in many microbiome workflows.

We could take an entire semester-long course on phylogenetics! This is because there is **a LOT** of evolutionary theory packed into phylogenetics. Some core themes include: 

1. **Evolutionary Concepts**
  - Common ancestry, homology vs. analogy
  - Gene trees vs. species trees
  - Molecular evolution basics
2. **Sequence Alignment**
  - Multiple sequence alignment (MSA)
  - Impact of alignment quality
  - Tools: MAFFT, MUSCLE, Clustal Omega
3. **Models of Molecular Evolution**
  - Nucleotide, amino acid, codon models
  - Model selection (e.g., GTR, HKY)
  - Rate heterogeneity (gamma, invariant sites)
4. **Tree-Building Methods**
  - Distance-based (NJ, UPGMA)
  - Maximum Parsimony
  - Maximum Likelihood (ML)
  - Bayesian Inference
5. **Tree Support & Comparison**
  - Bootstrapping
  - Posterior probabilities
  - Comparing tree topologies
6. **Phylogenetic Software**
  - FastTree, RAxML, IQ-TREE, MrBayes, BEAST, MEGA
  - Command-line vs. GUI tools
  - Workflow integration
7. **Phylogenomics**
  - Concatenated vs. coalescent approaches
  - Orthologs, paralogs, HGT
  - Species tree estimation
8. **Applications**
  - Ancestral state reconstruction
  - Biogeography
  - Microbial ecology (e.g., UniFrac)
  - Epidemiological phylogenetics
  9. **Challenges in Phylogenetics**
  - Incomplete lineage sorting
  - Long-branch attraction
  - Recombination and alignment uncertainty
  
For amplicon-based microbial community studies, FastTree2 is widely used because it strikes a balance between speed and phylogenetic accuracy, making it ideal for datasets with thousands of sequences. For whole-genome alignments or studies requiring branch support values, more computationally intensive tools like RAxML or IQ-TREE are preferred.

Here we will use FastTree2! 

# Setup 

## Timing of this script

Let's record how long this file took to run on the class server, which we will record at the end of the script. 

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set Seed 
```{r set-seed}
set.seed(238438)
```

## Load Packages 
```{r load-packages}
# Install phytools
#remotes::install_github("liamrevell/phytools")

# Install ggtree
#devtools::install_github("GuangchuangYu/ggtree")

# Now load them into the session 
pacman::p_load(phytools, ggtree, tidyverse, install = FALSE)
```

## 1. Load Data 

```{r load-physeq}
# Load physeq 
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq
```

# 2. Write ASV Fasta 

```{r asv-fasta}
# Pull out ASV sequences and ASV names
asv_seq_df <- 
  raw_preprocessed_physeq@tax_table %>% 
  data.frame() %>% 
  dplyr::select(ASV, ASVseqs)

#View(asv_seq_df)

# Add the ">" to make fasta header
asv_seq_df$ASV <- paste0(">", asv_seq_df$ASV)

#View(asv_seq_df)

# Create fasta object
asv_seq_fasta <- c(rbind(asv_seq_df$ASV, asv_seq_df$ASVseqs))
head(asv_seq_fasta)

# Write to a fasta file 
write(asv_seq_fasta, 
      file = "data/04_PhylogeneticTree/preprocessed_ASVs.fasta")

```

# 3. Run Alignment with MAAFT 

## What is multiple sequence alignment? 

**Multiple sequence alignment (MSA)** is the process of aligning three or more biological sequences (DNA, RNA, or protein) to identify regions of similarity. These conserved regions often indicate evolutionary relationships, structural or functional roles, and are critical for downstream analyses like phylogenetic tree building, motif discovery, or variant calling.

## Why MAFFT? 

MAFFT is fast and accurate multiple sequence alignment. MAFFT stands for **M**ultiple **A**lignment using **F**ast **F**ourier **T**ransform) is one of the most popular MSA tools due to its:

1. Speed – handles thousands of sequences quickly
2. Accuracy – multiple algorithms depending on dataset size
3. Scalability – works well for amplicon data (e.g., 16S rRNA) and large genomic datasets
4. Options – includes modes like `--auto`, `--globalpair`, `--linsi` for high accuracy

**Some quick notes before we MAFFT:**

Note that `engine.opts = '-l'` in our R chunk header name gives us our original terminal base environment. 

To test shell code via RMarkdown: 
- **Mac:** `command` + `option` + `enter` will send line-by-line code 
- **Windows:** `control` + `alt` + `enter`

```{r run-mafft, engine = 'bash', engine.opts = '-l'}
# Write bash code to run MAFFT
# First provide the path to MAFFT
export PATH=/programs/mafft/bin:$PATH

# Where am I? 
echo "The working directory is $PWD"

# Set a seed for consistency and reproducibility 
RANDOM=238428

# Now, actually run MAFFT
/programs/mafft/bin/mafft --auto data/04_PhylogeneticTree/preprocessed_ASVs.fasta > data/04_PhylogeneticTree/MAFFT_aligned_ASVs.fasta
```

# 4. Calculate the phylogenetic tree with FastTree2

[FastTree2](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0009490) is a widely used phylogenetic tree-building program that's particularly well-suited for microbial community analyses, especially when you're dealing with large datasets like 16S rRNA gene sequences from amplicon surveys. 

## What is FastTree2? 

It's **fast** because: 

- It uses approximate likelihoods rather than full ML computations.
- Applies heuristic neighbor-joining methods to reduce the number of topology changes it needs to test. 
- Optimizes only the top few candidate tree rearrangements rather than exhaustively testing all options.

This means it can handle alignments with thousands to tens of thousands of sequences in a few hours, where more rigorous methods might take days or weeks.

## Why is FastTree2 good for this application of microbial community analyses? 

1. **Scalability**: Amplicon datasets (*e.g.,* 16S, ITS) often have thousands of ASVs or OTUs—FastTree can process these without needing a supercomputer.
2. **Compatibility**: It's integrated into pipelines like QIIME 2, Phyloseq, and mothur, making it a seamless option for microbial ecologists.
3. **Speed vs. Accuracy Trade-off**: In microbial community analysis, you often care more about broad phylogenetic patterns (*e.g.,* for UniFrac) than about having the most perfect tree.
4. **Output Readiness**: FastTree produces newick files that can be directly used in downstream ecological analyses (*e.g.,* PCoA on UniFrac distances).

## When to NOT use FastTree? 

Do not use FastTree2 if:

- Publishing a high-resolution evolutionary analysis of closely related microbial strains or genomes.
- Doing bootstrapping for high-confidence clades. 
- Working with genome-scale alignments (e.g., core gene alignments from MAGs). 

In these cases, you can use tools like [IQ-TREE2](https://academic.oup.com/mbe/article/37/5/1530/5721363) or [RAxML](https://academic.oup.com/bioinformatics/article/30/9/1312/238053), which are more appropriate, as they provide more accurate trees and branch support values.

```{r run-fasttree2, engine = 'bash', engine.opts = '-l'}
# Where am I? 
echo "The working directory is $PWD"

# Load fasttree 
# Provide export path to fasttree 
export PATH=/programs/FastTree-2.1.11:$PATH

# Run Fasttree to generate phylogenetic tree 
# parameters: 
    # -nt = indicates it's a nucleotide alignment
    # -gtr = generalized time reversible substitution model 
    # -fastest speed up the model, reduce memory usage (recommended for datasets that have >50,000)
    # -log = output a log file 
    # input alignment file 
    # specify the output tree file 
FastTree -nt -gtr -fastest -log data/04_PhylogeneticTree/FastTree.log data/04_PhylogeneticTree/MAFFT_aligned_ASVs.fasta > data/04_PhylogeneticTree/ASVs_unrooted.tree
```

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```



